\input{pdfinfo}

\documentclass{article}

% Preamble
\input{preamble/preamble}

% Title
\title{Selected entrance examination problems at the Faculty of Mechanics and Mathematics of Moscow State University in 1986}
\author{}
\date{}


\begin{document}

\maketitle
\vspace{-0.4cm}
The level of difficulty of the problems is indicated by asterisks; in brackets - note from OCT [?].
\vspace{0.5cm}


\problem
Find all rational solutions to the equation
\begin{equation*}
  y = \sqrt{x^2 + \frac{11}{3}}
\end{equation*}

(A simplified analogue of this problem is in the book:
D.O. Shklyarsky, P.N. Chentsov, I.M. Yaglom
``Selected Problems and Theorems of Elementary Mathematics. Arithmetic and Algebra''
Problem No. 273.)


\problem
Using only a straightedge, construct a line paraller to two given parallel lines, passing through a given point.

(The problem is a viceroid, O. Shklyarsky, N.N. Chentsov, I.M. Yaglom ``Selected Problems and Theorems of Planimetry'' problem No. 65[?].
The solution to the problem given in the book is used in [?] as an auxiliary construction of problem 65a.)


\problem[{**}]
Compare
$\mspace{1.5mu}
 \dfrac{\log_3\paren{4} \log_3\paren{6} \log_3\paren{8} \cdots \log_3\paren{80}}
       {2 \log_3\paren{5} \log_3\paren{7} \log_3\paren{9} \cdots \log_3\paren{79}}
\mspace{1.5mu}$
and $1$.


\problem
Solve the equation
\begin{equation*}
  \paren{x - y}^2 + \paren{y - 2 \sqrt{x} + 2}^2 = \frac{1}{2}
\end{equation*}


\problem[{**}]
The base of the triangular pyramid $DABC$ is the equilateral triangle $ABC$.
Prove that if $\widehat{DAB} \cong \widehat{DBC} \cong \widehat{DCA}$,
then the pyramid is regular.

(The problem is in the problem book ``Kwanga'' (10, 1982), Problem 1770, with an asterisk, i.e., of greater difficulty.
One month is given to solve this section of the book.)


\problem[{*}]
Circle $O_1$ is drawn through three non-adjacent vertices of a cube.
Circle $O_2$ is inscribed in a face of the same cube that contains exactly one of these vertices.
Find the distance between the circles $O_1$ and $O_2$.

(The problem was part of the final round of the 1985 All-Soviet-Union Mathematical Olympiad (for 10th graders)).

Clearer formulation: \newline
Let $ABCD$ be a base of a cube.
Let $A' B' C' D'$ be the opposite base of that cube,
with $A'$ adjacent to $A$,
$B'$ adjacent to $B$,
$C'$ adjacent to $C$,
and $D'$ adjacent to $D$.
Detetrmine the distance between the circle inscribed in $ABCD$
and the circle circumscribed about $A$, $B'$ and $C$.


\problem
Let $ABC$ be a triangle.
Construct equilateral triangles $A B C'$, $B C A'$, $C A B'$
externally from $ABC$, using its sides as bases.
Prove that the lines $\overline{A A'}$, $\overline{B B'}$, $\overline{C C'}$ are concurrent.

(This is the fundamental lemma, essentially equivalent to the solution of problem 71[?]*
from the book by D.O. Shklyarsky, N.N. Chentsov, I.M. Yaglom ``Selected Problems and Theorems of Plane Geometry'',
a problem with an asterisk, i.e. of greater difficulty.) \newline
(V.V. Prosimov, Zatan N menuque, 9. 16)


\problem
Solve the following equation in rational numbers
\begin{equation*}
  \sqrt{2 \sqrt{3} - 3} = \sqrt{x \sqrt{3}} - \sqrt{y \sqrt{3}}
\end{equation*}


\problem
The median, angle bisector, and height from one vertex of a triangle
divide the corresponding angle into 4 equal parts.
Determine the angles of the triangle.

(The problem is in the book by D.O. Shklyarsky, N.N. Chentsov, I.M. Yaglom "Selected Problems and theorems in Plane Geometry'', problem [?] 113[?]).


\problem
Find all triangles that have an inscribed circle of radius $1$
and whose heights are integer.


\problem
Solve the equation
$x^2 - 5 = \sqrt{x + 5}$.

(A similar problem is in the book:
D.O. Shklyarsky, N.N. Chentsov, I.M. Yaglom
``Selected problems and theorems of elementary mathematics. Arithmetic and Algebra.''
Problem No. 239.)


\problem
Prove that if
$\dfrac{1}{\mspace{1mu} a + b + c \mspace{1mu}} = \dfrac{1}{a} + \dfrac{1}{b} + \dfrac{1}{c}$,
then, for any $n$, it holds that
$\dfrac{1}{\mspace{1mu} a^n + b^n + c^n \mspace{1mu}} = \dfrac{1}{a^n} + \dfrac{1}{b^n} + \dfrac{1}{c^n}$.

(The main auxiliary statement to this problem is in the book:
V.B. Lidsky, D.V. Ovsyannikov, A.N. Tulaikov, I. Shabunin
``Problems in Elementary Mathematics'', No. 262.
Within each subsection, the problems are ordered by their difficulty, increasing from the preface to the end.
Numbers in in the relevant subsection: 234 - 260.
In the written exam at [MITNX][?], where this problem was proposed, five hours were given to solve four problems.)


\problem[{*}]
Solve the equation
$x^4 + x^3 - 3 a x^2 - 2 a x + 2 a^2 = 0$,
for $x > 0$, $a > 1$.


\problem
Construct a tangent to the curve $y = \sqrt[3]{x}$
using only straightedge and compass. \newline
?[? Through a given point? or just one tangent will do, whichever it is?]

(The statement can be interpreted in two ways:
the first is that the problem is to be solved with the graph of the curve given;
in this case, the solution is simple.
The second is that the graph of the curve is not given,
only coordinate axes with unit segments on them are given.
Then, solving the problem for a point with abscissa $2$ is a equivalent to the problem of the doubling of a cube,
which is unsolvable with straightedge and compass.
Examiners used both interpretations of the condition.)


\problem[{\footnote{\label{merged_problems}Note: during the appeal it became known that the examiners considered these problems as one single problem. (The problems were offered on the same picture).}}]
A triangle $ABC$ (with $\overline{AB} \cong \overline{BC}$)
is inscribed in a circle so that the bisectors of the external angles at $B$ and $C$ intersect on the circle:
find the radius of the circle.

(The conditions are impossible.)


\problem[{\footref{merged_problems}}]
In a triangle $ABC$, the bisectors of the angles at $B$ and $C$ are drawn.
Does there exist a circle passing through the intersection point of these bisectors,
their bases [the vertices $B$ and $C$] and the vertex $A$?
If so, determine the type of triangle.


\problem[{\footref{merged_problems}}]
A circle is drawn through the bases of two angle bisectors
$\overline{BD}$ and $\overline{CK}$ of triangle $ABC$,
their point of intersection $P$,
and the third vertex $A$ of triangle $ABC$.
$\overline{DK} = 1$.
Determine $\overline{AD}$ and $\overline{AK}$.

(Insufficient data.
Solve the problem with the additional value $\widehat{ABC} = \alpha$.)


\end{document}
